The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-22 manuscript/rebuttal.Rmd

## Reviewer 3

> The paper is mostly well written and addresses an important topic: computational reproducibility and
> tools to achieve it. The authors describe five "pillars" under which various tools and techniques can
> be categorized. The paper is pretty comprehensive in what it covers and has useful supplementary
> material. My main issue with the paper is that it reiterates many things that have been covered in
> other papers. The authors cite 11 such papers. Many of the topics covered in submitted paper have
> already been covered well in other papers. I am most familiar with reference 13, which covers many of
> the same topics, although this paper goes into more detail on many issues. The authors could do more
> to differentiate their paper from previous ones and perhaps remove some topics that are already
> covered well elsewhere.

RESPONSE: XX

> Aside from that, I have listed below some relatively minor issues that, if addressed, would improve
> the paper. When I list page numbers, I am using the PDF page numbers rather than the numbers shown in
> the top-left corner of the manuscript.

@@

RESPONSE: True. We have amended the "Documentation" section to remove the unnecessary reiteration:

> As outlined above, one of the goals is to reduce the complexity of reproduction, in particular by
> minimising the number of commands required.

> 13. Section on version control: This section focuses mostly on using VC for software development
> (that is my interpretation).
> To be consistent with the introduction, it should focus more on data analyses.
> Although I use VC for analyses, I feel that simpler approaches are better in many cases.
> For example Dropbox and Google Drive provide some version-control and backup functionality and do
> not require the same level of knowledge as git.

RESPONSE: In accordance with this suggestion, we have rephrased some of the text to highlight the
utility of DVCS for data analysts.

> Although DVCSs come from the world of software engineering, data analysts can significantly
> benefit from their use in the life sciences [@ram2013; @blischak2016; @perez-riverol2016].

The bullet points following this passage outline the benefits of using DVCS.
While cloud storage like dropbox could protect against code loss, they cannot other five points cannot
be satisfied.

@@

> 14. Page 7: Many data analysts will not know what JupyterLab or VS Code are.
References are also needed.

RESPONSE: The phrase has been updated accordingly:

> `git` is typically used at the command line, however it is also incorporated into integrated
> development environments commonly used in bioinformatics including Rstudio, JupyterLab [68,69] and
> VS Code [70,71].

@@

> 15. Page 9, line 7: References are missing for these other tools.

TODO: Podman, Rkt or Apptainer/Singularity, Docker need refs.

@@

> 16. Page 9, lines 7-8: I believe you, but I am not aware of evidence that supports this claim.

TODO: "Docker remains the most widely used containerisation system for web development and research"

@@

> 17. The Biocontainers project should be mentioned.

RESPONSE: Biocontainers is mentioned in the last paragraph of the "Compute environment control" section
and has reference (Ref #91).

> 18. Page 10, lines 51-52: I disagree that the risk is small.
> There are many instances of using genomic data to identify individuals who have committed crimes.

RESPONSE: In light of this, we have changed the section in question:

> While deidentification of data has been widely used to safeguard participant privacy, this may not be
> suitable for genomics data due to the availability of existing public genetic data sets that heighten
> reidentification risks [105].
> For example, the surnames of some male participants can be inferred based on the Y chromosome
> variants seen in deidentified genomic data together with public geneology web sites [106].
> To foster the responsible reuse of sensitive genomic and health data, the Global Alliance for
> Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
> frameworks designed to protect personal data in a way that perserves reproducibility [107,108].

@@

> 19. Figures 2 and 4 are very similar to figures used in reference 13.

Figure 4 is adapted from the diagram on the Docker official website, which was originally published in
2015 or earlier (https://web.archive.org/web/20151023003338/https://www.docker.com/what-docker).
We have cited the original source.

Figure 2 is somewhat different from reference 13, as it shows an Rstudio window with windows for raw code
in text format beside the output document.
The figure in reference 13 shows a Jupyter notebook window, which has a different arrangement of code
and outputs.

> 20. Figure 5: I don't think it's really necessary to resummarize the FAIR principles.
> People can go to the source article for that.

RESPONSE: We acknowledge that it can sometimes be tiresome to reiterate things like the FAIR principles.
However summarising the key points succinctly will be in the interests of readers with limited time.
We note that the original description of the FAIR principles by Wilkinson and colleagues is ~5,000 words,
while this entire manuscript is ~6,500. We note that other reviewers did not insist on removing this
section.

@@

> 21. Page 11, line 13: I disagree that repositories like GEO and SRA are FAIR.
There are lots of problems with FAIRness in these repositories.

RESPONSE: While there may be problems with FAIRness with NCBI GEO and SRA, these repositries mostly
satisfy FAIR principles (eg: https://www.biorxiv.org/content/10.1101/739334v1.full.pdf).
From the perspective of a researcher sharing genomic data, they are "good enough" as compared to the
well known alternatives (eg: general purpose repository, institutional data storage, supplementary file,
dropbox, etc.).

@@

> 22. Page 11, lines 13-15: That's not true for some of these disciplines. Ecology has NEON, evo bio
has NCBI.

RESPONSE: We acknowledge that NCBI is a well known repository for evolutionary biology.
We have altered the section for clarity:

> The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find
> FAIR repositories that accepts data from other domains of study like ecology,
> physiology, molecular simulation, social sciences and computing [@Pampel2013-ep].

### 2023-08-22 manuscript/5pillars.Rmd
working on rev3

Computational researchers can benefit significantly by adopting DVCS into their routine
practices.
Below we outline some of the benefits.

@@

`git` is typically used at the command line, however it is also integrated into Rstudio, JupyterLab and
VS Code, and there are also several `git` clients with graphical interfaces which better allow inspection
of code changes.

Finally, although DVCSs come from the world of software engineering,
their use has become a standard in computational biology, as witnessed by the numerous guides
and best practices on the subject [@ram2013; @blischak2016; @perez-riverol2016].

@@

The field of genomics has a long history of data sharing @Brazma2001-yk, which is supported by
policy leadership by funding bodies
[@National_Health_and_Medical_Research_Council2019-yx;
@Office_of_The_Director_National_Institutes_of_Health2023-ss] and data sharing policies of journals
(eg [@Packer2018-ru; @Hanson2011-ub ; @Cheifet2020-wu]), although there are some challenges.
Some researchers would rather keep data for their private use.
Some may keep data private due to concerns over participant privacy.
This is a well-founded concern for identifiable data, however once identifying features are removed,
it is nearly impossible to re-identify genomic data @Venkatesaramani2021-tn unless the attackers have
pre-existing genomic data describing an individual or ancestry [@Schadt2012-md; @Gymrek2013-fz].
Therefore, sharing de-identified genomic data and metadata is likely to carry little risk to participant
privacy.
Still, much human genomic data and metadata resides in restricted databases due to the scope of
data sharing that was agreed to in the consent process.
For example, the consent form may only allow sharing of data to researchers investigating certain
diseases, which means a vetting process must be undertaken before sharing data with external
investigators.
In the interests of reuse and reproducibility, future studies could consider a consent form that allows
more broad sharing of de-identified data.

@@

High throughput biology has already built a number of data repositories for specific data types
conforming to FAIR principles (e.g.: GEO, SRA, ENA, PRIDE [@Edgar2002-hs; @Kodama2012-bn; @Deutsch2017-hc]).
However other domains lack well known specialized organized data repositories, such as evolutionary
biology, ecology, cell biology, molecular simulation, social sciences and computing.
The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find a
repository that accepts data from these fields [@Pampel2013-ep].
If no specialized repository exists, then a general-purpose repository such as
[Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
should be considered. Researchers should be aware, however, that these repositories are often not moderated, 
and that it is their responsibility to make their data FAIR, in particular by providing sufficient metadata.
To this aim, general recommendations have been proposed to enhance data FAIRness in ecology, but which
can also be applied to other disciplines [@Roche2015-aq].

>>>>

Although DVCSs come from the world of software engineering, data analysts can significantly
benefit from their use in the life sciences [@ram2013; @blischak2016; @perez-riverol2016].
Below we outline some of these advantages.

@@

`git` is typically used at the command line, however it is also incorporated into integrated development
environments commonly used in bioinformatics including Rstudio, JupyterLab
[@Kluyver2016jupyter;@noauthor_undated-hg] and VS Code [@noauthor_undated-fn ;@Extensions_undated-ln].
There are also several `git` clients with graphical interfaces which better allow inspection
of code changes (e.g.: @noauthor_undated-rh, @noauthor_undated-kj).

@@

The field of genomics has a long history of data sharing @Brazma2001-yk, which is supported by
policy leadership by funding bodies
[@National_Health_and_Medical_Research_Council2019-yx;
@Office_of_The_Director_National_Institutes_of_Health2023-ss] and data sharing policies of journals
(eg [@Packer2018-ru; @Hanson2011-ub ; @Cheifet2020-wu]), although there are some challenges.
Some researchers would rather keep data for their private use.
Some may keep data private due to concerns over participant privacy.

While deidentification of data has been widely used to safeguard participant privacy, this may not
be suitable for genomics data due to the availability of existing public genetic data sets that heighten
reidentification risks @Heeney2011-ri.
For example, the surnames of some male participants can be inferred based on the Y chromosome variants
seen in deidentified genomic data together with public geneology web sites @Gymrek2013-fz.
To foster the responsible reuse of sensitive genomic and health data, the Global Alliance for
Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
frameworks designed to protect personal data in a way that perserves reproducibility
[@Rehm2021-kt;@Thorogood2021-kt].

@@

High throughput biology has already built a number of data repositories for specific data types mostly
conforming to FAIR principles.
Repositories like GEO, SRA, ENA and PRIDE were developed in the aftermath of the draft human
genome assembly and have become well known to omics researchers
[@Edgar2002-hs; @Kodama2012-bn; Martens2005-wo]).
The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find
FAIR repositories that accepts data from other domains of study like ecology,
physiology, molecular simulation, social sciences and computing [@Pampel2013-ep].
If no specialized repository exists, then a general-purpose repository such as
[Dryad](https://datadryad.org/), [FigShare](https://figshare.com/), or [Zenodo](https://zenodo.org/)
should be considered. Researchers should be aware, however, that these repositories are often not moderated, 
and that it is their responsibility to make their data FAIR, in particular by providing sufficient metadata.
To this aim, general recommendations have been proposed to enhance data FAIRness in ecology, but which
can also be applied to other disciplines [@Roche2015-aq].
